Verification of Stencil Computations through Spatial Specifications

### 2016-07-07 paper.tex
eval section edits

\textbf{The Unified
  Model}~\cite{um}
is a weather forecasting and climate modelling tool developed by the
Met Office in the United Kingdom. It is used by research organisations
and meteorological services around the world. We use the development
branch (trunk) of the model. The code base in closed source but
institutional licenses are available for research purposes. The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model. We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.

@@

\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of tropospheric chemistry developed at Harvard and used by $\sim$70 universities and research institutions world-wide. We examined version 10-01.

@@

If the statement meets the stencil critera but the inference process
cannot determine an exact stencil for a statement it will instead
produce an upper (\term{atMost}) and lower (\term{atLeast}) bound. The
number of exact and inexact stencils are shown in columns three and
four of Table~\ref{tab:tickAssign}. Note that the total for ``Exact``
is higher than ``Actual`` because for some lines we produce multiple
specifications (perhaps for different array variables).
We see that our specification language is capable of precisely specifying the overwhelming majority of stencils. Most of the inexact stencils apply to some form of data copying. For example (taken from E3MG):
\begin{minted}{fortran}
!- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
!- stencil atMost, readOnce, (forward(depth=8, dim=2))*(reflexive(dim=1)) :: sfdt
ZZ1(I,J) = SFDT(I,J+8)
\end{minted}

@@

\textbf{All-reflexive} The vast majority (66\%) of stencils were
reflexive in every dimension of the array. These most often correspond
to applying point-wise transformations to data such as scaling
e.g. \begin{minted}{fortran}
ZZ1(I,K) = .000041868*ZZ1(I,K)
\end{minted}
Arrays with this class of stencil must be indexed by all the induction
variables used in the LHS without any offsets.

\textbf{Some-reflexive} stencils (24\%) consist of only reflexive terms but
leave some dimensions unspecified. Here are two examples:
\begin{minted}{fortran}
leafpools(:,age_class)=leafpool(:,1)
RVM(J,I2) = VRFO(ivar,J)
\end{minted}
where for the first example \mintinline{fortran}{leafpools} contains an
implicit loop, and for the second example \mintinline{fortran}{J} is
the induction variable. This stencil permits array indexing by a subset of the LHS induction variables but again without
any offsets.

\textbf{Single-action} stencils (6\%) consist of a single \term{backward},
\term{forward}, or \term{centered} region constant and some number of \term{reflexive}
region constants.
% /home/mrd45/src/corpus/geos-chem/GTMM/CasaRegridModule.F90:645
\begin{minted}{fortran}
do j=2,jm-ig
   do i=1,im
      != stencil readOnce, (backward(depth=1, dim=2))*(reflexive(dim=1)) :: dm, q
      al(i,j) = 0.5e+0_fp*(q(i,j-1)+q(i,j)) + r3*(dm(i,j-1) - dm(i,j))
   enddo
enddo
\end{minted}
This stencil requires a neighbourhood index including zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.

\textbf{Single-action-irreflexive} stencils (3\%) ) consist of a single
\term{backward}, \term{forward}, or \term{centered} irreflexive region constant
and some number of reflexive region constants.
% /home/mrd45/src/corpus/geos-chem/GeosRad/rrtmg_lw_rtrnmc.F90:355
\begin{minted}{fortran}
do lev = nlayers, 1, -1
    ! ...
    radld = radld - radld * (atrans(lev) + &
       efclfrac(lev,igc) * (1. - atrans(lev))) + &
       gassrc + cldfmc(igc,lev) * &
       (bbdtot * atot(lev) - gassrc)
    != stencil readOnce, (forward(depth=1, dim=2, irreflexive)) :: cldfmc
    drad(lev-1) = drad(lev-1) + radld
\end{minted}
Notice that, in this last example, the read from
\mintinline{fortran}{cldfmc} flows through onto the line with the
stencil. Stencils of this type required a neightbourhood index without a zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.

The remaining stencil shapes (less than 1\%) consisted primarily of multiple \term{forward}, \term{backward}, or \term{centered} dimensions. We note that only 0.2\% of stencils found required use of the union (\term{+}) operator. Despite this low frequency these stencils are arguably the most important since they represent more complicated code and thus greater potential for programming errors.

\subsection{Detecting programming errors}

\subsubsection{Example: 2-D Jacobi iteration}

One relatively basic example of a stencil program is two dimensional
Jacobi iteration that repeatedly goes through each cell in a matrix
and computes the average value of the four adjacent cells. The kernel
can be expressed in a line of Fortran code,
\begin{minted}{fortran}
  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
\end{minted}
and CamFort is able to infer a precise stencil specification for this
line,
\begin{minted}{fortran}
  != stencil reflexive(dim=1)*centered(depth=1, dim=2, irreflexive) + reflexive(dim=2)*centered(depth=1, dim=1, irreflexive) :: a
\end{minted}
which when inserted back into the program can be checked by CamFort as
well. We then tested the error-detecting ability of CamFort by
generating a family of 6,561 Jacobi iteration programs by perturbing
the indices used in the following manner: each access of the array was
templated as \mintinline{fortran}{a(i+ki,j+kj)} and then all possible
combinations of \mintinline{fortran}{ki} and \mintinline{fortran}{kj}
were drawn from the set $\{-1,0,1\}$. There are four uses of
\mintinline{fortran}{ki} and four of \mintinline{fortran}{kj}, hence
$3^8=6561$ possible perturbations of this kind. This particular setup
tests the class of errors in which the programmer typed an incorrect
number but at least got the ordering of the indices correct, although
CamFort would also detect a mixup of variables in this case as well.

CamFort checked all 6,561 programs and found that only 24 of them were
valid implementations of the stencil specification. This was the
expected result because there are $4!$ possible correct permutations
of the four terms in the kernel. All of the other programs were found
to violate the specification in some way. Another way to test the
programs is through visual examination of their output in graphical
format. Figure~\ref{fig:jacobi} shows some of the results of running
the Jacobi iteration programs, rendered into image files. All of the
correct variants emitted images indistinguishable from
Figure~\ref{fig:jacobicorrect}, while all of the incorrect variants
had wildly different image outputs. Both the ``visual test'' and a
byte-by-byte comparison showed results that completely concurred with
the results of the CamFort test.

@@

\subsubsection{Turbulence calculation}

>>>>

\textbf{The Unified
  Model}~\cite{um}
is a weather forecasting and climate modelling tool developed by the
Met Office in the United Kingdom. It is used by research organisations
and meteorological services around the world. We use the development
branch (trunk) of the model. The code base in closed source but
institutional licenses are available for research purposes. %The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model. We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.

@@

\textbf{GEOS-Chem}~\cite{geos-chem} is a three-dimensional model of
tropospheric chemistry developed at Harvard and used by $\sim$70
universities and research institutions world-wide. We use v.10-01.

@@

If the statement meets the stencil critera but the inference process
cannot determine an exact stencil for a statement it will instead
produce an upper (\term{atMost}) and lower (\term{atLeast}) bound. The
number of exact and inexact stencils are shown in columns three and
four of Table~\ref{tab:tickAssign}. Note that the total for ``Exact``
is higher than ``Actual`` because for some lines we produce multiple
specifications (perhaps for different array variables).
We see that our specification language is capable of precisely specifying the overwhelming majority of stencils. Most of the inexact stencils apply to some form of data copying. For example (taken from \textbf{E3MG}):
\begin{minted}{fortran}
!- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
!- stencil atMost, readOnce, (forward(depth=8, dim=2))*(reflexive(dim=1)) :: sfdt
ZZ1(I,J) = SFDT(I,J+8)
\end{minted}

@@

\textbf{All-reflexive} The vast majority (66\%) of specification are
reflexive in every dimension. These most often correspond
to point-wise transformations on data such as scaling, \eg{} from \textbf{E3MG}: \begin{minted}{fortran}
ZZ1(I,K) = .000041868*ZZ1(I,K)
\end{minted}
Arrays with this class of stencil are indexed by all the induction
variables used in the LHS without any offsets.

\textbf{Some-reflexive} specifications (24\%) consist of only reflexive terms but
leave some dimensions unspecified. These two examples are from
\textbf{GEOS-Chem} and \textbf{E3MG}: 
\begin{minted}{fortran}
leafpools(:,age_class)=leafpool(:,1)
RVM(J,I2) = VRFO(ivar,J)
\end{minted}
where for the first example \mintinline{fortran}{leafpools} contains an
implicit loop due to the range term \texttt{:}, and for the second \mintinline{fortran}{J} is
the induction variable. This stencil permits array indexing by a subset of the LHS induction variables but again without
any offsets.

\textbf{Single-action} specifications (6\%) consist of a single \term{backward},
\term{forward}, or \term{centered} region and some number of \term{reflexive}
region constants. For example, taken from \textbf{GEOS-Chem}: 
% /home/mrd45/src/corpus/geos-chem/GTMM/CasaRegridModule.F90:645
\begin{minted}[breakafter=*]{fortran}
do j=2,jm-ig
   do i=1,im
      != stencil readOnce, (backward(depth=1, dim=2)) *(reflexive(dim=1)) :: dm, q
      al(i,j) = 0.5e+0_fp*(q(i,j-1)+q(i,j)) + r3*(dm(i,j-1) - dm(i,j))
\end{minted}
% TODO: A BIT UNCLEAR
%This stencil requires a neighbourhood index including zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.
These kinds of specification require zero offsets in one dimension and 
and neighbourhood offsets (from 0 to $-1$ above) in the other.
The remaining dimensions must either have no offset or not pertain to
an induction variable.

\textbf{Single-action-irreflexive} specifications (3\%) consist of a single
\term{backward}, \term{forward}, or \term{centered} irreflexive region 
and some number of reflexive region constants, \eg{}, from \textbf{GEOS-Chem}:
% /home/mrd45/src/corpus/geos-chem/GeosRad/rrtmg_lw_rtrnmc.F90:355
\begin{minted}{fortran}
do lev = nlayers, 1, -1
    ! ...
    radld = radld - radld * (atrans(lev) + &
       efclfrac(lev,igc) * (1. - atrans(lev))) + &
       gassrc + cldfmc(igc,lev) * &
       (bbdtot * atot(lev) - gassrc)
    != stencil readOnce, forward(depth=1, dim=2, irreflexive) :: cldfmc
    drad(lev-1) = drad(lev-1) + radld
\end{minted}
Note that the read from 
\mintinline{fortran}{cldfmc} on line 5 flows to line 8. Stencils of
the above kind require a neightbourhood index without a zero-offset in
one dimension only. The remaining dimensions must either have no
offset or not pertain to an induction variable.

The remaining specification shapes (less than 1\%) consisted
primarily of multiple \term{forward}, \term{backward}, or
\term{centered} dimensions. We note that only $\sim$0.4\% of inferred
stencil specifications required use of the \term{+} operator. Despite their low frequency these stencils are arguably the most important since they represent more complicated code and thus greater potential for programming errors.

\subsection{Verification case study 1: 2-D Jacobi iteration}

One relatively basic example of a stencil computation is two-dimensional
Jacobi iteration that repeatedly goes through each cell in a matrix
and computes the average value of the four adjacent cells. The kernel
can be expressed in a line of Fortran code:
\begin{minted}{fortran}
  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
\end{minted}
and CamFort is able to infer a precise specification for this:
\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil reflexive(dim=1)*centered(depth=1,dim=2,irreflexive)+ reflexive(dim=2)*centered(depth=1,dim=1,irreflexive) :: a
\end{minted}
We then tested the error-detecting ability of our approach by 
generating a family of 6,561 Jacobi iteration programs by perturbing
the indices used in the following manner: each access of the array was
templated as \mintinline{fortran}{a(i+ki,j+kj)} and then all possible
combinations of \mintinline{fortran}{ki} and \mintinline{fortran}{kj}
were drawn from the set $\{-1,0,1\}$. There are four uses of
\mintinline{fortran}{ki} and four of \mintinline{fortran}{kj}, hence
$3^8=6561$ possible perturbations of this kind. This particular setup
tests the class of errors in which the programmer typed an incorrect
number but at least got the ordering of the indices correct, although
CamFort would also detect a mixup of variables in this case as well.

We assigned the specification inferred above to the generated Jacobi
stencils. CamFort checked all 6,561 programs against this spec and found that only 24 of them were
valid implementations of the stencil specification. This was the
expected result as there are $4!$ possible correct permutations
of the four terms in the kernel. All of the other programs were found
to violate the specification in some way. Another way to test the
programs is through visual examination of their output in graphical
format. Figure~\ref{fig:jacobi} shows some of the results of running
the Jacobi iteration programs, rendered into image files. All of the
correct variants emitted images indistinguishable from
Figure~\ref{fig:jacobicorrect}, while all of the incorrect variants
had wildly different image outputs. Both the ``visual test'' and a
byte-by-byte comparison showed results that completely concurred with
the results of the CamFort test.

@@

\subsection{Verification case study 2: Turbulence calculation}
